---
title: "Test current response extraction"
output: html_notebook
---



```{r include=FALSE, eval=FALSE}
rm(list = ls(all.names = TRUE))
```


## Read ABF files in R

To read some ABF files into R, we need to use the readABF package.


```{r setup}
library(data.table) # fast implementation of dataframe
library(tidyfast)   # fast manipulation of data.table
library(dtplyr)     # multithreaded data.table back-end for dplyr
library(tidyverse,  # everyone's first choice for data wrangling
        warn.conflicts = FALSE)
library(readABF)    # reads ABF files
library(checkmate)  # argument checking
library(tictoc)     # matlab-like execution timing
```


### Read ABF

```{r data1}
abf <- readABF('../../../dat/2017_03_09_5HT_run_0000.abf')
abf
```


```{r data2}
as.data.frame(abf, sweep = 1)
```
###### NOTE
...the convention of this library is to stamp the measurement with the end of
the first sample, but our convention is to put it at the beginning...


### Extract sweeps

We need to extract the FSCV data
```{r extraction}
# find the channel number with FSCV data (name can be FSCV or FCSV_1)
fscv_ch <- which(str_detect(abf$channelNames, fixed('FSCV')))

# if there are multiple FSCV channels, we need to split things up
if (length(fscv_ch) != 1) 
  stop(sprintf(
    'Found %d FSCV channels in ABF %s, expected 1',
    length(fscv_ch),
    abf$path))

# we'll use the standard 1000-sample voltammogram window, which depends on 
# the sweep size
sweep_size <- toString(dim(abf$data[[1]])[1])
sample_win <- switch(
  sweep_size,
  '1032' = 16:1015,
  '10000' = 160:1159,
  stop(sprintf(
    'Found unexpected sweep length %s in ABF %s',
    sweep_size,
    abf$path)))

# now that we know the channel number, we can extract FSCV data by mapping
# a function onto each element of the data list (i.e. sweep)
fscv_raw <- abf$data %>% 
  map(
    function(sweep) {
      return(sweep[sample_win , fscv_ch])
    })

# to get the data in a dataframe, we must first convert it to a matrix (sweep-per-row)
fscv_mat <- do.call(rbind, fscv_raw)

# we then need to give each sample column a name
colnames(fscv_mat) <- paste0('S', 1:(dim(fscv_mat)[2])) # S1, S2, etc.
```


### Melt to data.frame using dplyr
```{r extraction2.0}
#
# using dplyr, we can convert it to a dataframe with one sweep per row
#
melt_with_dplyr <- function(M) {
  as_tibble(M) %>% 
    mutate(  
      sweep = row_number()        # add sweep numbers
    ) %>% 
    pivot_longer(                 # tidy up so we have one sample per row...
      cols = -sweep,              # ...by converting all columns but 'sweep'...
      values_to = 'current',      # ...putting values in 'current' column...
      names_to = 'sample',        # ...putting colnames in 'sample' column...
      names_prefix = 'S',         # ...removing the initial "S"...
      names_transform = list(  
        sample = as.integer       # ...and converting result to integers.
      )
    ) %>% 
    arrange(sweep, sample) %>%   # sort by sweep and sample
    as.data.table()              # convert to data.table
}

tic()
fscv.0 <- melt_with_dplyr(fscv_mat)
toc()
```


### Melt to data.table using dtplyr
```{r extraction2.1}
#
# using dtplyr, we can try to do it faster...
#
melt_with_dtplyr <- function(M) {
  lazy_dt(M) %>%                  # convert to lazy data.table
    mutate(  
      sweep = row_number()        # add sweep numbers
    ) %>% 
    collect() %>%                 # collect results into a native data.table...
    dt_pivot_longer(              # tidy up so we have one sample per row...
      cols = -sweep,              # ...by converting all columns but 'sweep'...
      values_to = 'current',      # ...putting values in 'current' column...
      names_to = 'variable'       # ...and we'll ignore the column names
    ) %>% 
    lazy_dt() %>%                 # convert back to lazy data.table...
    group_by(sweep) %>%           # ...and group by sweep... 
    mutate(
      sample = row_number()       # ...add the sample numbers (pre-ordered)
    ) %>% 
    select(                       # order the output columns...
      sweep,
      sample, 
      current                     # ...excluding the extra column
    ) %>%         
    arrange(sweep, sample) %>%    # sort by sweep and sample
    as.data.table()               # convert to data.table
}

tic()
fscv.1 <- melt_with_dtplyr(fscv_mat)
toc()
```


### Melt to data.table using data.table native code
```{r extraction2.2}
#
# just for kicks, here's raw datatable code
#
melt_with_dt <- function(M) {
  check_matrix(M, 
               mode = 'double', 
               all.missing = FALSE, 
               min.rows = 150,
               ncols = 1000,
               null.ok = FALSE)
  
  DT <- as.data.table(M)[         # convert to data.table 
    , sweep := 1:.N               # add sweep numbers
  ] %>% 
    melt(                         # tidy up so we have one sample per row...
      id.vars = c('sweep'),       # ...by converting all columns but 'sweep'...
      value.name = 'current',     # ...putting values in 'current' column...
      variable.name = 'dummy'     # ...and we'll ignore the column names
    ) 
  
  setkey(DT, sweep)               # set a key on the sweep for easy grouping
  
  DT[ ,
     sample := 1:.N,              # add sample numbers...
     by = sweep                   # ...for each sweep
  ]
  
  DT[
    order(sweep, sample),         # order by sweep and sample...
    .(sweep, sample, current)     # ...and return "correct" column position
  ]
}


tic()
fscv.2 <- melt_with_dt(fscv_mat)
toc()
```


```{r equality}
# We have to skip checking of attributes because they keys are not set the same
# across the different methods

# - the dplyr method (fscv.0) does not create a key
key(fscv.0)

# - the dtplyr method (fscv.1) adds a key on column "sweep"
key(fscv.1)

# - the dt method (fscv.2) adds a key on column "sweep"
key(fscv.2)


all.equal(fscv.0, fscv.1, check.attributes = FALSE)
all.equal(fscv.1, fscv.2)
all.equal(fscv.2, fscv.0, check.attributes = FALSE) 
```


```{r benchmark}
# now we can test
results <- bench::mark(
  melt_with_dplyr(fscv_mat),
  melt_with_dtplyr(fscv_mat),
  melt_with_dt(fscv_mat),
  check     = FALSE,
  filter_gc = FALSE,
  min_iterations = 10
)

results %>% 
  mutate(expression = map_chr(expression, toString))
```
