---
title: "Comparison of overflow correction algorithms"
author: "Jason White"
date: "`r Sys.Date()`"
output: html_notebook
---

```{r include=FALSE, eval=FALSE}
rm(list = ls(all.names = TRUE))
```

### Libraries

```{r setup}
library(data.table)   # faster alternative to data.frame
library(tidyverse)    # everyone's first choice for data wrangling
library(readABF)      # reads ABF files
library(tictoc)       # matlab-like execution timing
library(checkmate)    # argument checking
library(matrixStats)  # for fast matrix ops

# include functions from preprocessing.R -- will be a library some day.
knitr::read_chunk('preprocessing.R')
```

### Load and preprocess data

```{r fn_read, echo=FALSE}
```

```{r fn_melt, echo=FALSE}
```

```{r}
tic()
abf1 <- read_abf('dat/ABF1.abf')
toc()

tic()
abf2 <- read_abf('dat/ABF2.abf')
toc()

# save some copies in case we need them
I1.raw <- copy(abf1$data)
I2.raw <- copy(abf2$data)
```

### Matlab function

##### (as of `r Sys.Date()`)

This function (not run here) is how the Matlab preprocessing pipeline corrects overflow.

It calculates the time difference of the recorded current $i_{rec}$, then looks for large negative values of $\Delta i_{rec}$, i.e. when $i_{rec}$ goes from $i_{max}$ to $i_{min}$ in one sample. These are marked as overflow startpoints. Similarly, it marks large positive values of $\Delta i_{rec}$ as overflow endpoints. It then sets $i_{rec}$ to 2000 nA between the start and end points.

```{octave, matlab_alg, eval=FALSE}
function V = correctOverflowClipping(V, dim, posthresh, negthresh)
    %CORRECTOVERFLOWCLIPPING Summary of this function goes here
    %   Detailed explanation goes here

    % default values
    if nargin < 4
        negthresh = -3500;
    end
    
    if nargin < 3
        posthresh = 3500;
    end
    
    if nargin < 2
        dim = 1;
    end
    
    % check arguments
    validateattributes(V, {'double'}, {'2d'}, 1);
    validateattributes(dim, {'double'}, {'scalar', 'positive', 'integer'}, 2);
    validateattributes(posthresh, {'double'}, {'scalar'}, 3);
    validateattributes(negthresh, {'double'}, {'scalar'}, 4);
    
    % use the diff to find overflow
    D = diff(V, 1, dim);
    
    % start of overflow is where diff is < negthresh
    S = find(D < negthresh);
    
    % end of overflow is where diff > posthresh
    E = find(D > posthresh);
    
    % check to see we have the same number of entries in S and E
    if numel(S) ~= numel(E)
        fprintf(' -- no undeflow correction (%d S, %d E) -- ', numel(S), numel(E));
        save('debug.mat', 'V', 'dim', 'posthresh', 'negthresh');
        return
    end
    
    % since S and E are in terms of diff, we need to identify the sweep
    % number to adjust the positions in terms of V
    sweep = floor(S ./ size(D, dim));
    
    % start of correction is one sample after S + sweep correction
    S = S + 1 + sweep;
    
    % end of correction is sample S + sweep correction
    E = E + sweep;
    
    % perform correction
    for ix = 1:numel(sweep)
        V(S(ix):E(ix)) = 2000;
    end
end
```

### Port above to R

```{r ported_fn}
correctOverflowClipping <- function(V, dimension = 1, posthresh = 3500, negthresh = -3500) {
  #CORRECTUNDERFLOWCLIPPING Summary of this function goes here
  #   Detailed explanation goes here
  
  # check arguments
  assert_matrix(V, mode = 'double', min.rows = 1, ncols = 1000)
  assert_integerish(dimension, len = 1, lower = 1, upper = 2)
  assert_number(posthresh, lower = 0, upper = 4000)
  assert_number(negthresh, lower = -4000, upper = 0)
  
  # use the diff to find overflow
  if (dimension == 1) {
    D <- rowDiffs(V)
  }
  else {
    D <- colDiffs(V)
  }
  
  # start of overflow is where diff is < negthresh
  S <- which(D < negthresh)
  
  # end of overflow is where diff > posthresh
  E <- which(D > posthresh)
  
  # check to see we have the same number of entries in S and E
  if (length(S) != length(E)) {
    stop(sprintf(
      ' -- no undeflow correction (%d S, %d E) -- ', length(S), length(E)))
  }
  
  # since S and E are in terms of diff, we need to identify the sweep
  # number to adjust the positions in terms of V
  sweep <- floor(S / dim(D)[dimension])
  
  # start of correction is one sample after S + sweep correction
  S <- S + 1 + sweep
  
  # end of correction is sample S + sweep correction
  E <- E + sweep
  
  # perform correction
  for (ix in seq_along(sweep)) {
    V[S[ix]:E[ix]] <- 2000
  }
  
  V
}
```

### Try it

```{r fig1a, fig.width=6}
test.matlab <- function(M) {
  fixed.matlab <- tryCatch(
    correctOverflowClipping(M),
    error = function(e) e
  )
  
  if (is.matrix(fixed.matlab)) {
    print(
      fixed.matlab %>% 
        as_melted_data_table() %>% 
        as_tibble() %>% 
        filter(sweep == 100) %>% 
        ggplot(aes(x = sample, y = current)) +
        geom_line() +
        geom_point(size = 1/2, color = 'blue') +
        labs(title = 'Matlab method') +
        theme_minimal()
    )
  } else {
    print(glue::glue('ERROR: {fixed.matlab$message}'))
  }
  
  invisible(fixed.matlab)
}

test.matlab(abf1$data)
```

The matlab algorithm breaks for this case, but the R algorithm works fine...

### R algorithm

```{r fn_fix}
```

```{r fig2a, fig.width=6}
test.r <- function(M){
  # first, melt the matrix into a data.table
  fixed.r <- M %>% 
    as_melted_data_table()
  
  fixed.r <- tryCatch(
    fix_overflow(fixed.r),
    error = function(e) e
  )
  
  if (is.data.table(fixed.r)) {
    print(
      fixed.r %>% 
        as_tibble() %>% 
        filter(sweep == 100) %>% 
        ggplot(aes(x = sample, y = current)) +
        geom_line() +
        geom_point(size = 1/2, color = 'blue') +
        labs(title = 'R method') +
        theme_minimal()
    )
  } else {
    print(glue::glue('ERROR: {fixed.matlab$message}'))
  }
  
  invisible(fixed.r)
}

test.r(abf1$data)
```

### Here's what the original looks like

```{r fig2c, fig.width=6}
I1.raw %>% 
  as_melted_data_table() %>% 
  as_tibble() %>% 
  filter(sweep == 100) %>% 
  ggplot(aes(x = sample, y = current)) +
  geom_line() +
  geom_point(size = 1/2, color = 'blue') +
  labs(title = 'Raw data') +
  theme_minimal()
```

### Here's a case where both methods complete, but the answer is different

```{r fig3, fig.width=6, fig.show="hold", out.width="50%"}
res.matlab <- test.matlab(abf2$data) %>% 
  as_melted_data_table() # melt for comparison

res.r <- test.r(abf2$data)[ 
  , 
  .(sweep, sample, current)  # remove extra columns for comparison
]

# compare
all.equal(res.r, res.matlab)
```

### Zoom in to see what's different

```{r fig4, fig.width=6}
cmp2 <- merge(
  x = res.r,
  y = res.matlab, 
  by = c('sweep', 'sample')
)

cmp <- merge(
  x = cmp2,
  y = as_melted_data_table(I2.raw),
  by = c('sweep', 'sample')
)

cmp %>% 
  as_tibble() %>% 
  filter(sweep == 100) %>% 
  ggplot(aes(x = sample)) +
  geom_line(aes(y = current, color = '(none)')) +
  geom_point(aes(y = current, color = '(none)'), size = 1/2) +
  geom_line(aes(y = current.y, color = 'Matlab')) +
  geom_point(aes(y = current.y, color = 'Matlab'), size = 1/2) +
  geom_line(aes(y = current.x, color = 'R')) +
  geom_point(aes(y = current.x, color = 'R'), size = 1/2) +
  coord_cartesian(xlim = c(100, 500)) +
  scale_color_manual(
    name = 'method', 
    values = c('Matlab' = 'red', 'R' = 'black', '(none)' = 'lightblue')) +
  labs(title = 'Overflow correction comparison of first half', y = 'current') +
  theme_minimal()
```

We can see by this that the Matlab method is overeager when it comes to detecting overflow.

------------------------------------------------------------------------

###### Notebook produced with the these R libraries

::: {style="width:33%; height:auto; margin-left:0%; margin-right:auto"}
```{r versions, echo=FALSE, rows.print=50}
ver <- paste0(version$major, '.', version$minor)

rbind(
  tribble(
    ~library, ~version, 
    'R',       str_remove(version$version.string, 'R version '),
    'RStudio', toString(RStudio.Version()$version)
  ),
  tibble(library = .packages()) %>% 
    mutate(version = map_chr(library, ~ toString(packageVersion(.x)))) %>% 
    arrange(library) %>% 
    filter(library %in% c(
      'data.table',
      'tidyverse',
      'readABF',
      'tictoc',
      'checkmate',
      'matrixStats'
    )) 
) %>% 
  knitr::kable()
```
:::
